Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-13 submission/manuscript.Rmd
add females to discussion

In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the adult _Kdm5c_-KO hippocampus and amygdala. In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain. Deeper characterization of testis-enriched DEGs revealed they were not somatic testis genes, but germline genes, thus demonstrating KDM5C's crucial role in establishing the soma-germline boundary. Given the majority of tissue-enriched DEGs are testis and ovary genes with no known brain functions, it is currently unknown if they impair _Kdm5c_-KO brain function and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression. However, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus. Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010].

We found a substantial portion of genes upregulated in _Kdm5c_-KO brain are germline genes with no known fucntions within the brain. Distinguishing the germline from the soma is a key feature of multicellularity and sexual reproduction[@devlinPanmetazoanViewGermlinesoma2023]. Previous work characterizing chromatin regualtors that silence germ cell-specific transcription has predominatly focused on their repression of key marker genes in embryonic stem cells (ESCs), such as _Dazl_ and _Ddx4_[@endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. To characterize KDM5C's role in germline gene repression at a genome-wide level throughout life, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015]. This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) which groups of germline genes are directly and indirectly regulated by KDM5C. This list of germline-enriched genes serves as a resource for future studies to systematically assess soma-germline dysregulation.

RNA-seq of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis. _In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into somatic tissues[@bardotMouseGastrulationCoordination2020]. However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011]. Therefore, misexpression of germline genes in EpiLCs might suggest _Kdm5c_-KO ESCs are progressing beyond EpiLC differentiation and instead becoming PGCLCs. Yet, _Kdm5c_-KO EpiLCs properly express primed pluripotency genes, including _Otx2_ - an epiblast stem cell marker that is known to repress differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Furthermore, we observed no difference in cellular morphology during _Kdm5c_-KO ESC to EpiLC differentiation. Proper EpiLC differentiation, together with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.

- XX vs XY
- While many germline genes have roles in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. Many of the germline genes characterized in the male _Kdm5c_-KO brain have sperm-specific functions. It is currently unclear if sperm genes are also dysregulated in female _Kdm5c_ mutants, or if they would instead misexpress egg-biased genes.

>>>>

In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the adult _Kdm5c_-KO hippocampus and amygdala. In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain. Given the majority of tissue-enriched DEGs have no known brain functions, it is currently unknown if they impair _Kdm5c_-KO neurodevelopment and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression. However, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus. Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010], however further investigation is required to determine its impact upon _Kdm5c_-KO phenotypes.

Deeper characterization of testis-enriched DEGs revealed they were not somatic testis genes but instead germline genes, thus demonstrating KDM5C's crucial role in establishing the soma-germline boundary. Distinguishing the germline from the soma is a key feature of multicellularity and sexual reproduction[@devlinPanmetazoanViewGermlinesoma2023]. Previous work characterizing repressors of germ cell-specific transcription has predominatly focused on their repression of key marker genes in embryonic stem cells (ESCs), such as _Dazl_ and _Ddx4_[@endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. To characterize KDM5C's role in germline gene repression throughout life at a genome-wide level, we curated a list of germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015]. This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, 3) the impact of germ cell sex upon germline gene misexpression, and 4) which groups of germline genes are directly and indirectly regulated by KDM5C. This list of germline-enriched genes serves as a resource for future studies to systematically assess soma-germline dysregulation.

Analysis of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis. _In vivo_, germline genes are typically fully decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into somatic tissues[@bardotMouseGastrulationCoordination2020]. However, a small subset of epiblast stem cells will reactivate germline genes to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011]. Thus, misexpression of germline genes in _Kdm5c_-KO EpiLCs might suggest ESCs are progressing beyond EpiLC differentiation and instead becoming PGCLCs. Yet, _Kdm5c_-KO EpiLCs properly express genes for primed pluripotency, including _Otx2_ - an epiblast stem cell marker that is known to repress differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Furthermore, we observed no difference in cellular morphology during _Kdm5c_-KO ESC to EpiLC differentiation. Proper EpiLC differentiation, coupled with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.

While many genes involved in early germline development and meiosis are common between males and females, some germline genes display sex-biased expression or perform functions unique to eggs and sperm. We found both males and females expressed both sperm and egg biased genes, indicating the mechanisms underlying germline gene repression are independent of germ cell sex. However, we found organismal sex did greatly influence the manifestation of germline gene dysregulation, as _Kdm5c_-KO females had over double the number of germline-enriched DEGs compared to male EpiLCs. Knockout of KDM5C in females is embryonic lethal, likely due to impaired X chromosome inactivation (XCI) [@samantaActivationXistEvolutionarily2022]. XCI defects could explain why _Kdm5c_-KO females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, ectopic germline transcripts, including those unique to females, were not significantly biased towards the X chromsome. Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, females heterozygous for _Kdm5c_ loss also had over double the number of germline DEGs than males, even though their XCI is largely preserved and their expression of KDM5C should be roughly equivalent to that of wild-type males. Altogether, this suggests female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity. Future studies investigating sex differences in germline gene repression are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators like KDM5C or a general feature of female cells.
